---
draft: true
format: pdf
---
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(nycflights13)
library(palmerpenguins)
moorea_coral <- read_csv("../../eod/data/moorea_coral.csv")
moorea_fish <- read_csv("../../eod/data/moorea_fish.csv")
```

<!-- Problem 1 -->

> I added a column called `Year` to the `moorea_coral` data frame but it's not there in the output.

```{r}
#| results: hold

moorea_coral |> mutate(Year = ym(Date))
glimpse(moorea_coral)
```

{{< pagebreak >}}

<!-- Problem 2 -->

> Why can't I select the departure day-of-week column?

```{r}
#| error: true
#| results: hold

flights |>
  mutate(
    departure = make_datetime(year, month, day, hour, minute),
    dep_wday = wday(departure, label = TRUE)
  ) |>
  select(departure, dep-wday)
```

{{< pagebreak >}}

<!-- Problem 3 -->

> Why can't I select the departure day-of-week column?

```{r}
#| error: true
#| results: hold

flights |>
  mutate(
    departure = make_date(year, month, day,hour, minute)
    dep_wday = wday(departure, label = TRUE)
  ) |>
  select(departure, dep_wday)
```

{{< pagebreak >}}

<!-- Problem 4 -->

> Collapsing the factors didn't work.

```{r}
#| results: hold

gss_cat |>
  mutate(
    pardyid = fct_collapse(
      partyid,
      "R" = c("Strong republican", "Not str republican"),
      "I" = c("Ind,near rep", "Independent", "Ind,near dem"),
      "D" = c("Not str democrat", "Strong democrat"),
      "O" = c("No answer", "Don't know", "Other party")
    )
  ) |>
  count(partyid)
```

{{< pagebreak >}}

<!-- Problem 5 -->

> What do you mean body mass not found, it's literally in the data frame.

```{r}
#| error: true
#| results: hold

penguins |>
  filter(!is.na(sex)) |>
  mutate(body_mass_kg = body_mass_g / 1000)
  summarize(
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),
    .by = c(species, sex)
  )
```

{{< pagebreak >}}

<!-- Problem 6 -->

> I summarized max_bill_length, why can't I plot it?

```{r}
#| error: true
#| results: hold

penguins |>
  summarize(
    max_bill_length = max(bill_length_mm, na.rm = TRUE),
    max_bill_depth = max(bill_depth_mm, na.rm = TRUE),
    .by = species
  ) |>
  ggplot(
    data = penguins,
    mapping = aes(x = max_bill_length, y = max_bill_depth)
  ) +
  geom_point()
```

{{< pagebreak >}}

<!-- Problem 7 -->

> Biomass is in grams, and I **know** there were piscivorous fish bigger than 100 g at every site. So why is my result an empty data frame?

```{r}
#| results: hold

moorea_fish |>
  filter(
    Coarse_Trophic == "Piscivore primarily",
    Biomass > 100
  ) |>
  count(Site)
```

{{< pagebreak >}}

<!-- Problem 8 -->

> Are you kidding me I just want to summarize these dang fish WHAT IS GOING ON

```{r}
#| error: true
#| results: hold

moorea_fish |>
  filter(Site == "LTER_1") |>
  summarize(
    mean_biomass = mean(Biomass, na.rm = TRUE),
    .by = Year, Habitat, Family
  )
```

