---
format: 
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme: 
      - ../../meds-slides-styles.scss
execute: 
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"} 

[EDS 221 Day 3 PM]{.custom-title}

[*Loops*]{.custom-subtitle}

<hr class="hr-teal">

[August 12^th^, 2026]{.custom-subtitle3}

---

## {#loops-repeat data-menu-title="Loops let you repeat yourself"} 

[Loops let you repeat yourself]{.slide-title}

<hr>

[For example, print each value in a vector]{.body-text-m}

```{r}
corals <- c("Porites", "Pocillopora", "Acropora")
for (c in corals) {
  print(c)
}
```

---

## {#loops-repeat2 data-menu-title="Loops let you repeat yourself"} 

[Loops let you repeat yourself]{.slide-title}

<hr>

[It's common to loop over the indices of a vector.]{.body-text-m}

```{r}
corals <- c("Porites", "Pocillopora", "Acropora")
for (i in 1:length(corals)) {
  print(corals[i])
}
```

---

## {#fibonacci data-menu-title="What does this loop make?"} 

[What does this loop make?]{.slide-title}

<hr>

[Let's break down this loop.]{.body-text-m}

```{r}
fibonacci <- integer(6)
fibonacci[1:2] <- 1
for (i in 3:length(fibonacci)) {
  fibonacci[i] <- fibonacci[i - 1] + fibonacci[i - 2]
}
fibonacci
```

---

## {#loop-parts data-menu-title="Parts of a loop"} 

[Parts of a loop]{.slide-title}

<hr>

[A loop has three main parts]{.body-text-m}


::: {.columns}

::: {.column width="30%"}

- **Iterator**
  `i`
- **Sequence**
  `3:length(fibonacci)`
- **Body**
  `{ ... }`

:::

::: {.column width="70%"}

```{r}
fibonacci <- integer(6)
fibonacci[1:2] <- 1
for (i in 3:length(fibonacci)) {
  fibonacci[i] <- fibonacci[i - 1] + fibonacci[i - 2]
}
fibonacci
```

:::

:::

---

## {#iterator data-menu-title="Parts of a loop"} 

[Parts of a loop]{.slide-title}

<hr>

[A loop has three main parts]{.body-text-m}


::: {.columns}

::: {.column width="30%" .body-text-m}

The **iterator** changes value each time through the loop.

:::

::: {.column width="70%"}

```{r}
fibonacci <- integer(6)
fibonacci[1:2] <- 1
for (i in 3:length(fibonacci)) {
  fibonacci[i] <- fibonacci[i - 1] + fibonacci[i - 2]
}
fibonacci
```

:::

:::

---

## {#sequence data-menu-title="Parts of a loop"} 

[Parts of a loop]{.slide-title}

<hr>

[A loop has three main parts]{.body-text-m}

::: {.columns}

::: {.column width="30%" .body-text-m}

The **sequence** is what the iterator takes on values from.

:::

::: {.column width="70%"}

```{r}
fibonacci <- integer(6)
fibonacci[1:2] <- 1
for (i in 3:length(fibonacci)) {
  fibonacci[i] <- fibonacci[i - 1] + fibonacci[i - 2]
}
fibonacci
```

:::

:::

---

## {#body data-menu-title="Parts of a loop"} 

[Parts of a loop]{.slide-title}

<hr>

[A loop has three main parts]{.body-text-m}


::: {.columns}

::: {.column width="30%" .body-text-m}

The **body** is what the loop repeatedly does.

:::

::: {.column width="70%"}

```{r}
fibonacci <- integer(6)
fibonacci[1:2] <- 1
for (i in 3:length(fibonacci)) {
  fibonacci[i] <- fibonacci[i - 1] + fibonacci[i - 2]
}
fibonacci
```

:::

:::

---