---
format: 
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme: 
      - ../../meds-slides-styles.scss
execute: 
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"} 

[EDS 221 Day 4 PM]{.custom-title}

[*Vectorized functions*]{.custom-subtitle}

<hr class="hr-teal">

[August 13^th^, 2026]{.custom-subtitle3}

---

## {#vectorization data-menu-title="Vectorization"} 

[Vectorization]{.slide-title}

<hr>

**Vectorized** functions and operators work on entire vectors at once, rather than element by element.

Using vectorization often improves code performance and clarity.

```r
squares <- c(1, 4, 9, 16, 25)

# Loop
roots <- double(length(squares))
for (i in 1:length(roots)) {
  roots[i] <- sqrt(roots)
}

# Vectorized
roots <- sqrt(squares)
```

---

## {#functions-operators data-menu-title="Functions and operators"} 

[Functions and operators]{.slide-title}

<hr>

Both functions and operators can be vectorized.

```{r}
a <- 1:4
b <- 4:1

a + b
as.character(a)
```

---

## {#review- data-menu-title="Review"} 

[Review]{.slide-title}

<hr>

The `round()` function rounds numbers.

```{r}
round(pi, digits = 2)
```

Consider this expression:

```r
round(seq(0, 2, by = 1/2) * pi, digits = 2)
```

Using only pencil and paper, answer the following questions.


::: {.callout-tip}

Solve sub-expressions first to build up to the complete solution.

:::


1. What will the output be?
2. Which parameter is `round()` _vectorized_ over?

---
