---
format:
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - ../../meds-slides-styles.scss
execute:
  echo: true
  fig-width: 4
  fig-height: 3
---

## {#title-slide data-menu-title="Title Slide" background="#053660"}

[EDS 221 Day 6 PM]{.custom-title}

[*Grammar of graphics*]{.custom-subtitle}

<hr class="hr-teal">

[August 17^th^, 2026]{.custom-subtitle3}

---

## {#grammar-overview data-menu-title="The grammar of graphics"}

[The grammar of graphics]{.slide-title}

<hr>

[`ggplot2` builds a figure by layering independent pieces on top of each other, instead of picking one big "chart type."]{.body-text-m}

::: {.columns}

::: {.column}

**_data_**

The data frame you're plotting.

:::

::: {.column}

**_mapping_**

Which columns control which visual properties (`aes()`).

:::

::: {.column}

**_geometry_**

The shape used to represent the data (`geom_*()`).

:::

::: {.column}

**_scale_**

How the mapping shows up on the figure (`scale_*()`).

:::

:::

---

## {#ggplot-data data-menu-title="Start with data"}

[Geometries: start with data]{.slide-title}

<hr>

[`ggplot()` sets up a plot for a data frame — with nothing mapped yet, there's nothing to draw.]{.body-text-m}

```{r}
#| message: false
library(tidyverse)
library(palmerpenguins)

ggplot(data = penguins)
```

---

## {#ggplot-mapping .smaller data-menu-title="Add a mapping"}

[Geometries: add a mapping]{.slide-title}

<hr>

[`aes()` maps columns to axes. This sets up the grid, but still no geometry has been added.]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
)
```

---

## {#ggplot-geom-point .smaller data-menu-title="Add a geometry"}

[Geometries: add a geometry]{.slide-title}

<hr>

[A `geom_*()` layer draws the mapped data. `geom_point()` draws one point per row.]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

---

## {#geom-choices .smaller data-menu-title="Choosing a geometry"}

[Geometries: choosing a geometry]{.slide-title}

<hr>

[The geometry you choose depends on the kind of relationship you want to show.]{.body-text-m}

::: {.columns}

::: {.column}

[One numeric variable → `geom_histogram()`]{.body-text-s}

```{r}
#| fig-height: 3
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram()
```

:::

::: {.column}

[One categorical variable → `geom_bar()`]{.body-text-s}

```{r}
#| fig-height: 3
ggplot(penguins, aes(x = island)) +
  geom_bar()
```

:::

:::

---

## {#geom-boxplot .smaller data-menu-title="Comparing groups"}

[Geometries: comparing groups]{.slide-title}

<hr>

[A numeric variable by a categorical variable → `geom_boxplot()` shows the distribution within each group.]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = species, y = body_mass_g)
) +
  geom_boxplot()
```

---

## {#aesthetics-overview data-menu-title="Aesthetics"}

[Aesthetics]{.slide-title}

<hr>

[`aes()` isn't just for `x` and `y` — it can map columns to other visual properties too.]{.body-text-m}

- `x`, `y` — position
- `color` — outline / line color
- `fill` — interior color
- `shape` — point symbol

---

## {#aes-color-fill .smaller data-menu-title="color vs. fill"}

[Aesthetics: color vs. fill]{.slide-title}

<hr>

::: {.columns}

::: {.column}

[`color` — outlines]{.body-text-s}

```{r}
#| fig-height: 3.5
ggplot(
  penguins,
  aes(x = species, y = body_mass_g, color = sex)
) +
  geom_boxplot()
```

:::

::: {.column}

[`fill` — interiors]{.body-text-s}

```{r}
#| fig-height: 3.5
ggplot(
  penguins,
  aes(x = species, y = body_mass_g, fill = sex)
) +
  geom_boxplot()
```

:::

:::

---

## {#aes-shape .smaller data-menu-title="shape"}

[Aesthetics: shape]{.slide-title}

<hr>

[`shape` gives each category of a point geometry a different symbol — useful when color alone isn't enough (e.g., grayscale printing).]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g, y = flipper_length_mm, shape = species)
) +
  geom_point()
```

---

## {#aes-color-vs-shape .smaller data-menu-title="Combining aesthetics"}

[Aesthetics: combining color and shape]{.slide-title}

<hr>

[Mapping `species` to both `shape` and `color` encodes the same grouping two ways, making it easier to read.]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = body_mass_g,
    y = flipper_length_mm,
    shape = species,
    color = species
  )
) +
  geom_point()
```

---

## {#scales-overview data-menu-title="Scales"}

[Scales]{.slide-title}

<hr>

[Every mapped aesthetic has a **scale** controlling how data values become visuals — an axis, a legend, a color ramp. `scale_*()` functions let you customize them: titles, breaks, and the values themselves.]{.body-text-m}

---

## {#scales-axis .smaller data-menu-title="Renaming axes"}

[Scales: renaming and rebreaking axes]{.slide-title}

<hr>

[`scale_x_continuous()` and `scale_y_continuous()` control the x- and y-axis scales — here, a cleaner title and fewer tick marks (`n.breaks`).]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g, y = flipper_length_mm, shape = species)
) +
  geom_point() +
  scale_x_continuous(name = "Body mass (g)", n.breaks = 2) +
  scale_y_continuous(name = "Flipper length (mm)", n.breaks = 4)
```

---

## {#scales-manual .smaller data-menu-title="Manual scales"}

[Scales: choosing values manually]{.slide-title}

<hr>

[`scale_shape_manual()` lets you assign the exact shape (or color, or fill) for each category, instead of accepting the default.]{.body-text-m}

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g, y = flipper_length_mm, shape = species)
) +
  geom_point() +
  scale_shape_manual(
    name = "Species",
    values = c(Adelie = "square", Chinstrap = "diamond", Gentoo = "circle")
  )
```

---
