---
format:
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - ../../meds-slides-styles.scss
execute:
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"}

[EDS 221 Day 6 AM]{.custom-title}

[*Data wrangling*]{.custom-subtitle}

<hr class="hr-teal">

[August 17^th^, 2026]{.custom-subtitle3}

---

## {#meet-the-data .smaller data-menu-title="Meet the data: penguins"}

[Meet the data: penguins]{.slide-title}

<hr>

[Today we'll wrangle a data frame of penguin measurements from the Palmer Archipelago, Antarctica.]{.body-text-m}

```{r}
#| message: false
library(tidyverse)
library(palmerpenguins)

penguins
```

---

## {#glimpse .smaller data-menu-title="glimpse()"}

[Getting an overview: glimpse()]{.slide-title}

<hr>

[`glimpse()` shows every column, its type, and a preview of its values — a fast way to see what you're working with.]{.body-text-m}

```{r}
glimpse(penguins)
```

---

## {#filter-rows .smaller data-menu-title="filter(): working with rows"}

[Working with rows: filter()]{.slide-title}

<hr>

[`filter()` keeps only the rows that match a condition. The first argument is the data frame; the rest are conditions that must be `TRUE`.]{.body-text-m}

```{r}
filter(penguins, species == "Chinstrap")
```

---

## {#filter-conditions .smaller data-menu-title="filter(): combining conditions"}

[filter(): combining conditions]{.slide-title}

<hr>

::: {.columns}

::: {.column}

[`,` or `&` — **both** must be true]{.body-text-m}

```{r}
filter(
  penguins,
  sex == "female",
  flipper_length_mm > 190
)
```

:::

::: {.column}

[`|` — **either** can be true]{.body-text-m}

```{r}
filter(
  penguins,
  species == "Chinstrap" |
    body_mass_g < 3000
)
```

:::

:::

[This `|` is not the same as the `||` from last week's conditionals — `|` compares element-by-element across an entire column.]{.body-text-s}

---

## {#arrange-rows .smaller data-menu-title="arrange(): working with rows"}

[Working with rows: arrange()]{.slide-title}

<hr>

[`arrange()` sorts rows by one or more columns, smallest to largest by default. Wrap a column in `desc()` to sort largest to smallest.]{.body-text-m}

::: {.columns}

::: {.column}

```{r}
arrange(penguins, bill_depth_mm)
```

:::

::: {.column}

```{r}
arrange(penguins, desc(bill_depth_mm))
```

:::

:::

---

## {#mutate-columns .smaller data-menu-title="mutate(): working with columns"}

[Working with columns: mutate()]{.slide-title}

<hr>

[`mutate()` adds a new column, computed from existing ones. The argument name becomes the new column name.]{.body-text-m}

```{r}
mutate(penguins, body_mass_kg = body_mass_g / 1000)
```

---

## {#mutate-multiple .smaller data-menu-title="mutate(): multiple columns"}

[mutate(): multiple new columns]{.slide-title}

<hr>

[You can create several columns in one call — later columns can even use ones you just created.]{.body-text-m}

```{r}
mutate(
  penguins,
  body_mass_kg = body_mass_g / 1000,
  bill_length_cm = bill_length_mm / 10
)
```

---

## {#select-columns .smaller data-menu-title="select(): working with columns"}

[Working with columns: select()]{.slide-title}

<hr>

[`select()` keeps only the columns you name, in the order you name them — everything else is dropped.]{.body-text-m}

```{r}
select(penguins, species, island, body_mass_g)
```

---

## {#select-helpers .smaller data-menu-title="select(): helpers"}

[select(): helpers]{.slide-title}

<hr>

::: {.columns}

::: {.column}

[`:` — a range of columns]{.body-text-m}

```{r}
select(penguins, species:bill_length_mm)
```

:::

::: {.column}

[`starts_with()` — match column names]{.body-text-m}

_There are lots of these helper functions!_

```{r}
select(penguins, starts_with("bill"))
```

:::

:::

---

## {#nested-calls .smaller data-menu-title="The problem with nesting"}

[The problem with nesting]{.slide-title}

<hr>

[Wrangling usually takes several steps. Nesting function calls works, but it reads inside-out and gets hard to follow.]{.body-text-m}

```{r}
filter(
  mutate(
    penguins,
    body_mass_kg = body_mass_g / 1000
  ),
  body_mass_kg < 2.8
)
```

---

## {#the-pipe .smaller data-menu-title="The pipe |>"}

[The pipe: |>]{.slide-title}

<hr>

[`|>` takes what's on its left and passes it in as the **first argument** of the function on its right. Steps now read top-to-bottom, in the order they happen.]{.body-text-m}

```{r}
penguins |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  filter(body_mass_kg < 2.8)
```

---

## {#summarize data-menu-title="summarize(): summarizing groups"}

[Summarizing groups: summarize()]{.slide-title}

<hr>

[`summarize()` collapses many rows into one summary value. Most summary functions need `na.rm = TRUE` to ignore missing values.]{.body-text-m}

```{r}
penguins |>
  summarize(avg_size = mean(body_mass_g, na.rm = TRUE))
```

---

## {#summarize-by data-menu-title="summarize(): grouped with .by"}

[summarize(): grouped with .by]{.slide-title}

<hr>

[The `.by` argument runs the summary separately for each group, returning one row per group.]{.body-text-m}

```{r}
penguins |>
  summarize(
    avg_size = mean(body_mass_g, na.rm = TRUE),
    biggest_bill = max(bill_length_mm, na.rm = TRUE),
    .by = species
  )
```

---

## {#memorize data-menu-title="What to memorize"}

[What to memorize]{.slide-title}

<hr>

[Data wrangling with the tidyverse involves far more functions than you've seen so far. You should memorize a few fundamentals now, then pick up the rest as you gain experience.]{.body-text-m}


::: {.columns}

::: {.column}

### Rows

- `filter()`
- `arrange()`

### Columns

- `select()`
- `mutate()`

:::

::: {.column}

### Groups

- `summarize()`
- `n()`
- The `.by` parameter

:::

:::
