---
format:
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - ../../meds-slides-styles.scss
execute:
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"}

[EDS 221 Day 7 PM]{.custom-title}

[*Reshaping data*]{.custom-subtitle}

<hr class="hr-teal">

[August 18^th^, 2026]{.custom-subtitle3}

---

## {#tidy-data data-menu-title="Tidy data"}

[Tidy data]{.slide-title}

<hr>

[A data frame is **tidy** when:]{.body-text-m}

- Every **variable** is a column
- Every **observation** is a row
- Every **value** is a cell

[The same values can be arranged into many different table shapes — only one of them is tidy. *Reshaping* moves values between rows and columns without changing what they mean.]{.body-text-s}

---

## {#meet-the-data-pm .smaller data-menu-title="Meet the data: penguins, summarized"}

[Meet the data: penguins, summarized]{.slide-title}

<hr>

[Average body mass by species and sex — one row per group. This is **long** data: each combination of `species` and `sex` gets its own row.]{.body-text-m}

```{r}
#| message: false
library(tidyverse)
library(palmerpenguins)

mass_by_sex <- penguins |>
  filter(!is.na(sex)) |>
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE), .by = c(species, sex))

mass_by_sex
```

---

## {#pivot-wider .smaller data-menu-title="pivot_wider(): making data wider"}

[pivot_wider(): making data wider]{.slide-title}

<hr>

[`pivot_wider()` spreads one column's values into new columns. `names_from` supplies the new column names; `values_from` supplies the cells.]{.body-text-m}

```{r}
mass_wide <- mass_by_sex |>
  pivot_wider(names_from = sex, values_from = mean_mass)

mass_wide
```

---

## {#pivot-wider-why .smaller data-menu-title="pivot_wider(): why reshape?"}

[pivot_wider(): why reshape?]{.slide-title}

<hr>

[Wide data makes it easy to compare values that were in *different rows* — now they're in different columns of the *same* row.]{.body-text-m}

```{r}
mass_wide |>
  mutate(diff = male - female)
```

[This comparison would need a join if `male` and `female` were still in separate rows.]{.body-text-s}

---

## {#pivot-longer .smaller data-menu-title="pivot_longer(): making data longer"}

[pivot_longer(): making data longer]{.slide-title}

<hr>

[`pivot_longer()` is the reverse: it gathers columns into rows. `cols` says which columns to gather; `names_to`/`values_to` name where they land.]{.body-text-m}

```{r}
mass_wide |>
  pivot_longer(
    cols = c(female, male),
    names_to = "sex",
    values_to = "mean_mass"
  )
```

---

## {#pivot-longer-billboard .smaller data-menu-title="pivot_longer(): a messier example"}

[pivot_longer(): a messier example]{.slide-title}

<hr>

[`billboard` tracks each song's chart rank in a separate column per week — `wk1` through `wk76`. That's 76 columns encoding one variable: *week*.]{.body-text-m}

```{r}
billboard |> select(artist, track, wk1, wk2, wk3)
```

---

## {#pivot-longer-billboard2 .smaller data-menu-title="pivot_longer(): starts_with()"}

[pivot_longer(): selecting with starts_with()]{.slide-title}

<hr>

[`cols` accepts the same tidyselect helpers as `select()`. `values_drop_na` removes weeks a song hadn't charted yet or had already dropped off.]{.body-text-m}

```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

---

## {#selecting-cols data-menu-title="Selecting columns to pivot"}

[Selecting columns to pivot]{.slide-title}

<hr>

[`cols` can be written several ways — pick whichever most clearly states your intent.]{.body-text-m}

- `c(female, male)` — name them explicitly
- `female:male` — a range of adjacent columns
- `starts_with("wk")` — match a naming pattern
- `!year` — everything **except** a column

---

## {#pm-memorize data-menu-title="What to memorize"}

[What to memorize]{.slide-title}

<hr>

[The same values can live in a wide or a long table — reshaping picks whichever shape fits the next step of your analysis.]{.body-text-m}

::: {.columns}

::: {.column}

### Wider

- `pivot_wider()`
- `names_from`, `values_from`
- Good for comparing values across a row

:::

::: {.column}

### Longer

- `pivot_longer()`
- `cols`, `names_to`, `values_to`
- Good for plotting and grouped summaries

:::

:::
