---
format:
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - ../../meds-slides-styles.scss
execute:
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"}

[EDS 221 Day 7 AM]{.custom-title}

[*Joining data*]{.custom-subtitle}

<hr class="hr-teal">

[August 18^th^, 2026]{.custom-subtitle3}

---

## {#meet-the-data .smaller data-menu-title="Meet the data: nycflights13"}

[Meet the data: nycflights13]{.slide-title}

<hr>

[Instead of one giant table, `nycflights13` spreads information about 2013 NYC flights across **five** related data frames.]{.body-text-m}

```{r}
#| message: false
library(tidyverse)
library(nycflights13)

flights
```

---

## {#relational-data data-menu-title="Relational data"}

[Relational data]{.slide-title}

<hr>

[A *relational dataset* stores information across multiple tables instead of one, connected by shared columns called **keys**.]{.body-text-m}

- `flights` — one row per flight
- `airlines` — one row per carrier
- `airports` — one row per airport
- `planes` — one row per physical aircraft
- `weather` — one row per airport, per hour

[Splitting data this way avoids repeating an airline's full name on every one of its thousands of flights.]{.body-text-s}

---

## {#primary-key .smaller data-menu-title="Primary keys"}

[Keys: primary key]{.slide-title}

<hr>

[A **primary key** uniquely identifies each row *in its own table*. Every plane appears exactly once in `planes`, identified by `tailnum`.]{.body-text-m}

```{r}
planes |>
  count(tailnum) |>
  filter(n > 1)
```

[Zero rows out means every `tailnum` is unique — `tailnum` is a valid primary key for `planes`.]{.body-text-s}

---

## {#foreign-key .smaller data-menu-title="Foreign keys"}

[Keys: foreign key]{.slide-title}

<hr>

[A **foreign key** refers to a primary key in *another* table. `tailnum` also appears in `flights` — but a single plane makes many flights.]{.body-text-m}

```{r}
flights |>
  count(tailnum) |>
  filter(n > 1)
```

[Rows here mean `tailnum` repeats in `flights` — it's a foreign key there, not a primary key.]{.body-text-s}

---

## {#inner-join .smaller data-menu-title="inner_join(): matching columns"}

[inner_join(): matching columns]{.slide-title}

<hr>

[`inner_join()` adds columns from a second table by matching rows on a shared key. `join_by()` says which column to match on.]{.body-text-m}

```{r}
flights |>
  inner_join(airlines, join_by(carrier)) |>
  select(flight, carrier, name)
```

---

## {#inner-join-rename .smaller data-menu-title="inner_join(): different column names"}

[inner_join(): different column names]{.slide-title}

<hr>

[When the key has different names in each table, connect them with `==` inside `join_by()`. Here `dest` in `flights` matches `faa` in `airports`.]{.body-text-m}

```{r}
flights |>
  inner_join(airports, join_by(dest == faa)) |>
  select(flight, dest, name)
```

---

## {#join-row-counts .smaller data-menu-title="What happens to row count"}

[What happens to row count]{.slide-title}

<hr>

[`inner_join()` keeps only rows with a match in **both** tables. Some flights use planes missing from `planes`, so they drop out.]{.body-text-m}

```{r}
nrow(flights)

flights |>
  inner_join(planes, join_by(tailnum)) |>
  nrow()
```

[Always check row counts before and after a join — an unexpected change usually means a key wasn't as unique as you thought.]{.body-text-s}

---

## {#join-family data-menu-title="The join family"}

[The join family]{.slide-title}

<hr>

[`inner_join()` is one of several **mutating joins** that add columns from a second table. They differ in which unmatched rows they keep.]{.body-text-m}

- `inner_join()` — keep rows that match in **both** tables
- `left_join()` — keep **all** rows from the left table
- `right_join()` — keep **all** rows from the right table
- `full_join()` — keep **all** rows from **either** table

[`left_join()` is the most common — it never drops rows from the table you started with.]{.body-text-s}

---

## {#filtering-joins .smaller data-menu-title="Filtering joins"}

[Filtering joins]{.slide-title}

<hr>

[**Filtering joins** use a second table to keep or drop rows from the first — without adding any new columns.]{.body-text-m}

::: {.columns}

::: {.column}

[`semi_join()` — keep rows *with* a match]{.body-text-s}

```{r}
carriers_with_data <- head(airlines, 5)

flights |>
  semi_join(carriers_with_data, join_by(carrier)) |>
  distinct(carrier)
```

:::

::: {.column}

[`anti_join()` — keep rows *without* a match]{.body-text-s}

```{r}
flights |>
  anti_join(carriers_with_data, join_by(carrier)) |>
  distinct(carrier)
```

:::

:::

---

## {#am-memorize data-menu-title="What to memorize"}

[What to memorize]{.slide-title}

<hr>

[Joining lets you work with data split across multiple tables instead of duplicating information in one giant sheet.]{.body-text-m}

::: {.columns}

::: {.column}

### Keys

- Primary key: unique in its own table
- Foreign key: refers to another table's primary key

:::

::: {.column}

### Joins

- `inner_join()`, `left_join()`, `full_join()`
- `join_by()` — and `join_by(x == y)` when names differ
- `semi_join()` / `anti_join()` to filter, not add columns

:::

:::
