---
format:
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - ../../meds-slides-styles.scss
execute:
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"}

[EDS 221 Day 8 PM]{.custom-title}

[*Dates and times*]{.custom-subtitle}

<hr class="hr-teal">

[August 19^th^, 2026]{.custom-subtitle3}

---

## {#why-dates-tricky data-menu-title="Why dates are tricky"}

[Why dates are tricky]{.slide-title}

<hr>

[Dates look like plain text, but they aren't — `"01/02/2017"` could mean January 2^nd^ or February 1^st^ depending on where you're reading from.]{.body-text-m}

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(lubridate)

class("2017-01-31")
```

[A character string has no concept of "one day later." A proper **date** type does.]{.body-text-s}

---

## {#parsing-dates .smaller data-menu-title="Parsing dates: ymd(), mdy(), dmy()"}

[Parsing dates: ymd(), mdy(), dmy()]{.slide-title}

<hr>

[**lubridate**'s parsing functions are named for the *order* the year, month, and day appear in your text — they all return the same kind of date.]{.body-text-m}

```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```

---

## {#parsing-datetimes .smaller data-menu-title="Parsing datetimes: ymd_hms()"}

[Parsing datetimes: ymd_hms()]{.slide-title}

<hr>

[Add `_hms`, `_hm`, or `_h` to also capture a time of day. Unspecified time components default to zero.]{.body-text-m}

```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
```

---

## {#extracting-components .smaller data-menu-title="Extracting components"}

[Extracting components]{.slide-title}

<hr>

[Once something is a date, you can pull out any piece of it — useful for grouping and filtering.]{.body-text-m}

```{r}
release_date <- ymd("2026-08-19")

year(release_date)
month(release_date, label = TRUE)
wday(release_date, label = TRUE)
```

---

## {#date-arithmetic .smaller data-menu-title="Date arithmetic"}

[Date arithmetic]{.slide-title}

<hr>

[Dates support arithmetic directly. `days()`, `weeks()`, and friends add a fixed span of calendar time.]{.body-text-m}

```{r}
start <- ymd("2026-08-19")

start + days(1)
start + weeks(2)
ymd("2026-12-25") - start
```

[The last line returns a `difftime` — the number of days between two dates.]{.body-text-s}

---

## {#pm-memorize data-menu-title="What to memorize"}

[What to memorize]{.slide-title}

<hr>

[Dates and datetimes are a distinct type from text — parse early, and let lubridate handle the arithmetic.]{.body-text-m}

::: {.columns}

::: {.column}

### Creating

- `ymd()` (and permutations)
- `ymd_hms()` (and permutations)

:::

::: {.column}

### Components

- `year()`, `month()`, `day()`, `wday()`
- `hour()`, `minute()`, `second()`

:::

:::
