---
format:
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - ../../meds-slides-styles.scss
execute:
  echo: true
---

## {#title-slide data-menu-title="Title Slide" background="#053660"}

[EDS 221 Day 8 AM]{.custom-title}

[*Factors*]{.custom-subtitle}

<hr class="hr-teal">

[August 19^th^, 2026]{.custom-subtitle3}

---

## {#meet-the-data .smaller data-menu-title="Meet the data: gss_cat"}

[Meet the data: gss_cat]{.slide-title}

<hr>

[`gss_cat` (from **forcats**) is a sample of the General Social Survey. Several columns store categorical responses.]{.body-text-m}

```{r}
#| message: false
#| warning: false
library(tidyverse)

gss_cat
```

---

## {#what-is-a-factor data-menu-title="What is a factor?"}

[What is a factor?]{.slide-title}

<hr>

[A **factor** stores categorical data as a set of predefined **levels** — the possible values a variable can take.]{.body-text-m}

```{r}
levels(gss_cat$rincome)
```

[Under the hood, R stores the values as integers and keeps a lookup table of level labels. This is what makes factors compact *and* orderable.]{.body-text-s}

---

## {#why-factors .smaller data-menu-title="Why not just use character strings?"}

[Why not just use character strings?]{.slide-title}

<hr>

[Character vectors sort **alphabetically**. Factors let you control the order directly, which matters for plots and tables.]{.body-text-m}

```{r}
gss_cat |>
  mutate(rincome = as.character(rincome)) |>
  ggplot(aes(x = rincome)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```

[Notice the income bins are in no meaningful order — alphabetical, not low-to-high.]{.body-text-s}

---

## {#fct-relevel .smaller data-menu-title="fct_relevel(): manual order"}

[fct_relevel(): manual order]{.slide-title}

<hr>

[`fct_relevel()` moves specific levels to the front, in the order you list them.]{.body-text-m}

```{r}
gss_cat |>
  mutate(marital = fct_relevel(marital, "Never married", "Married")) |>
  pull(marital) |>
  levels()
```

---

## {#fct-infreq .smaller data-menu-title="fct_infreq(): order by frequency"}

[fct_infreq(): order by frequency]{.slide-title}

<hr>

[`fct_infreq()` reorders levels from most to least common — useful whenever "biggest first" is the natural reading order.]{.body-text-m}

```{r}
gss_cat |>
  mutate(marital = fct_infreq(marital)) |>
  ggplot(aes(x = marital)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```

---

## {#fct-reorder .smaller data-menu-title="fct_reorder(): order by another variable"}

[fct_reorder(): order by another variable]{.slide-title}

<hr>

[`fct_reorder()` sorts levels by a *summary statistic* of another variable — e.g., average TV hours by religion.]{.body-text-m}

```{r}
relig_summary <- gss_cat |>
  filter(!is.na(tvhours)) |>
  summarize(tv_hours = mean(tvhours), .by = relig)

relig_summary |>
  mutate(relig = fct_reorder(relig, tv_hours)) |>
  ggplot(aes(x = tv_hours, y = relig)) +
  geom_point()
```

---

## {#fct-recode data-menu-title="fct_recode(): renaming levels"}

[fct_recode(): renaming levels]{.slide-title}

<hr>

[`fct_recode()` changes level *labels* without touching the underlying data or order. New name on the left, old name on the right.]{.body-text-m}

```{r}
gss_cat |>
  mutate(
    partyid = fct_recode(
      partyid,
      "Republican, strong" = "Strong republican",
      "Republican, weak" = "Not str republican"
    )
  ) |>
  count(partyid)
```

---

## {#fct-lump .smaller data-menu-title="fct_lump(): collapsing rare levels"}

[fct_lump(): collapsing rare levels]{.slide-title}

<hr>

[`fct_lump_n()` keeps the *n* most common levels and lumps the rest into `"Other"` — handy when a factor has too many rare categories to plot cleanly.]{.body-text-m}

```{r}
gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 5)) |>
  count(relig, sort = TRUE)
```

---

## {#am-memorize data-menu-title="What to memorize"}

[What to memorize]{.slide-title}

<hr>

[Factors store categorical data with explicit, controllable **levels** instead of relying on alphabetical order.]{.body-text-m}

::: {.columns}

::: {.column}

### Inspecting

- `levels()`

:::

::: {.column}

### Reordering / recoding

- `fct_reorder()`, `fct_recode()`

:::

:::
